Designing Types for R, Empirically

### 2020-09-10 paper/main.tex
minor edits and merge

@@

The next step, is shown in Table~\ref{fig:singlefreq}, the table provides a
breakdown of types occuring in arguments without a union. Scalar, class and
vector are the most common type categories. The shaded rows correspond to
polymorphic types. Removing those gives us \CountMonomorphicPositions
(\PercMonomorphicPositions) monomorphic positions in a corpus of
\CountPositions parameters. With close to 70\% of monomorphic argument or
return values, it is fair to say that even a simple type language provides
significant benefits.

@@

\paragraph{Structs}
While experimenting with various design, we consider adding a struct type to
capture lists with named elements that can be accessed with the \code{$}
operators. We ended discarding those types as they grew large and were often
only representative of the example data being manipulated.  Consider
function \code{cv.model}, its argument \code{x} is observed to be of
\CLASS{\texttt{aov}, \texttt{lm}} or \CLASS{\texttt{lm}}).  Internally,
linear models are represented as lists with named elements.  The polution is
illustrated by the lines after the funtion definition.  They load an example
data set and test the function \code{cv.model}.  The
\code{data(sweetpotato)} expression load a sample data set. The fields of
\code{sweetpotato} will be recorded when \code{cv.model} is called.

@@

\paragraph{Objects}
While we record classes, our analysis does not deal with method dispatch.  R
has multiple, disparate, object systems, called S3, S4, and R5.  The
\code{class} attribute is used by these systems to dispatch methods. S3 does
single dispatch, S4 does multiple dispatch and R5 supports impertative
objects.  The mechanics of S4 dispatch are more complex than for S3, and
users can define their own class hierarchies that we would need to
incorporate in our type analysis and contract checking frameworks.  We found
limited use of S4 during our analysis. Coming up with a type system that
accounts for all of these factors and consolidates multiple
object-orientation frameworks in a single language design is an interesting
problem in and of itself, one we leave for future work.

@@

>>>>

\AT{TODO: Softer intro to R.
People always get confused about c() and stuff like that.
We have some space, we can take it.}

@@

The next step, is shown in Table~\ref{fig:singlefreq}, the table provides a
breakdown of types occuring in arguments without a union. Scalar, class and
vector are the most common type categories. The shaded rows correspond to
polymorphic types. Removing those gives us \CountMonomorphicPositions
(\PercMonomorphicPositions) monomorphic positions in a corpus of
\CountPositions parameters. With close to 65\% of monomorphic argument or
return values, it is fair to say that even a simple type language provides
significant benefits.

We would like to point out the number of occurrences of the singleton {\tt null} type, and discuss why we treat it as an {\it any}-type.
In R, it is common for programmers to include default values for arguments, and in many cases this value happens to be {\tt NULL}.
Our type analysis will report a {\tt null} type for these arguments if they are never passed a value during testing.
Thus, we interpret these as cases where we cannot be sure of the argument.

@@

\paragraph{Structs}
While experimenting with various design, we consider adding a struct type to
capture lists with named elements that can be accessed with the \code{$}
operators. We ended up discarding those types as they grew large and were often
only representative of the example data being manipulated.  Consider
function \code{cv.model}, its argument \code{x} is observed to be of
\CLASS{\texttt{aov}, \texttt{lm}} or \CLASS{\texttt{lm}}.  Internally,
linear models are represented as lists with named elements.  The polution is
illustrated by the lines after the funtion definition.  They load an example
data set and test the function \code{cv.model}.  The
\code{data(sweetpotato)} expression load a sample data set. The fields of
\code{sweetpotato} will be recorded when \code{cv.model} is called.

@@

\paragraph{Objects}
While we record classes, our analysis does not deal with method dispatch.  R
has multiple, disparate, object systems, called S3, S4, and R5.  The
\code{class} attribute is used by these systems to dispatch methods. S3 does
single dispatch, S4 does multiple dispatch and R5 supports imperative
objects.  The mechanics of S4 dispatch are more complex than for S3, and
users can define their own class hierarchies that we would need to
incorporate in our type analysis and contract checking frameworks.  We found
limited use of S4 during our analysis. Coming up with a type system that
accounts for all of these factors and consolidates multiple
object-orientation frameworks in a single language design is an interesting
problem in and of itself, one we leave for future work.

@@

\AT{Do we want to expand on this evaluation?
Or just expand in a more general section later.}

 %
 %
 \subsubsection{Kaggle}

 \AT{TODO: Fill in and discuss Kaggle results.}

 To further validate our inferred types, we repeated the experiment discussed in this section on Kaggle code.
 By-and-large, we saw no meaningful difference in the two data sets, with the contract assertion failure patterns being repeated from the reverse dependencies.

 %
 %
 \subsubsection{Working Failed Assertions into Types}

 \AT{TODO: compare with types that we get when looking at reverse dependencies, and discuss implications.}

Throughout this section, we have seen instances where the reverse dependencies of a package exercise its functions in a way that was not accounted for in the package's own test and  example code.
These do not always represent errors, and may represent cases of oversight by package writers.

If we {\it include} running a package's reverse dependencies in our type inference process, we will obtain different types for certain functions.
In doing so, we observe \ldots
